Measuring Interaction Design \\before Building the System: a Model-Based Approach

### 2016-09-06 paper.tex
details
added copyright piece

\documentclass{sigchi}

@@

% Load basic packages
\usepackage{balance}  % to better equalize the last page
\usepackage{graphics} % for EPS, load graphicx instead 
%\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage{times}    % comment if you want LaTeX's default font
\usepackage[pdftex]{hyperref}
% \usepackage{url}      % llt: nicely formatted URLs
\usepackage{color}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{ccicons}
\usepackage{todonotes}
\newcommand{\mytodo}[1]{\todo[inline]{#1}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\migtool}{MIGtool}
\usepackage{soul}
\usepackage{paralist}
\usepackage{listings}

@@

% \author{Giorgio Brajnik\inst{1,2} \and Simon Harper\inst{1}}
% \institute{\textsuperscript{1}Computer Science School\\
% University of Manchester, UK\\
% \email{simon.harper@manchester.ac.uk}\\
% and\\
% \inst{2}Dipartimento di Matematica e Informatica\\
% Universit\`a di Udine, Italy\\
% \email{brajnik@uniud.it}
% }

@@

\keywords{Experimental; Evaluation; Statecharts: UML; Metrics; Testing.}

@@

\section{Introduction}

% \mytodo{ Subcommittee: Technology, Systems, and Engineering

% People we should pay particular attention to:
% Caroline Appert
% Conversy, Stéphane
% Nebeling, Michael

% Keyword we need to pay attention to:
% This subcommittee will focus on technology, systems and engineering
% contributions that enable, improve, or advance interaction. This will
% include software and hardware technologies and systems that enable and
% demonstrate novel interactive capabilities, as well as languages,
% methods and tools for \hl{CONSTRUCTION AND ENGINEERING} of interactive
% systems. Engineering contributions should clearly demonstrate how they
% address interactive systems concerns such as, for example,
% scalability, reliability, interoperability, testing, and
% performance. Systems and technology contributions will be judged by
% their technical innovation and/or ability to connect, SIMPLIFY or
% enrich interactions, for example in intelligent interfaces and
% mobile/ubiquitous computing.
% }
% % 
%\end{quote}

We present an approach that allows a designer to assess interaction
design qualities such as efficiency, error-proneness and
recovery from errors. Key importance is given to the ability to
\begin{inparaenum}[(a)]
\item understand how supportive a user interface (UI) is with respect
  to user efficiency;
\item understand how prone the UI is to user navigation
  errors;
\item understand how recoverable the UI is from those errors; and,
\item perform objective quantitative comparison of different designs
  to support construction and engineering of interactive systems.
\end{inparaenum}
All this can be done before developing prototypes of the UI, similarly
to what is claimed and done in~\cite{silva14}: ``... the analysis of solutions in
the early phases of development is crucial as their results can then
be iteratively injected in the development cycle''.

@@

Developing good UIs for web or mobile applications is a
complex and expensive endeavor. One reason is the combination of
devices, interaction modalities and workflows that need to be
supported.
%
Adopting Usage Centered Development practices is effective, as
is following established design principles~\cite{constantine99}. In
particular, one of the most effective technique is early prototyping
to explore part of the five-dimensional fidelity prototyping
space~\cite{mccurdy06,buxton07}. This is particularly effective when
paired with usability investigations based on user testing or
heuristic evaluations~\cite{preece02,rubin08}.

@@

Even worse, several questions cannot be easily
answered. Given one or more potential designs and some usage
scenarios, interesting questions include: ``How many different
routes can be followed by the user to carry out the scenario?'',
``Which are the shortest ones?'', ``If a user makes a mistake, would
he or she be able to recover?'', ``How many steps would the recovery
require?''. When designing and evaluating
embedded UIs (such as when dealing with plane's
cockpits~\cite{conversy07}), other relevant questions include
``How would the above properties change if we add a certain a
widget?'', or ``... if we replace a widget with another?''.  At the
moment, these straightforward questions are quite complicated to
answer. In fact, they require developing  prototypes, inspecting
them, manual tracking of which screens and widgets are used at which
stage, followed by a systematic manual analysis.

This should not be the case, however, because answers to these
questions could be automatically computed. In this way a system could
provide important insights to a designer, and support assessments of
potential user flexibility, user efficiency, error proneness, ability
to recover, compactness and consistency of a design.

Our approach is based on UML statechart models of UIs which are
automatically processed to produce \emph{interaction graphs}. These
graphs are then used to specify interaction scenarios and to unfold
the possible interaction paths (called \emph{execution traces}) that
are compatible with the specific scenarios being considered. Traces
are fed to graph-theoretic computations which produce a
dashboard with different results. Except for development of models and
specification of the desired scenarios, which have to be done
manually, the other steps are totally automatic; models of a design
(such as the ones shown below) can be developed in a matter of a
couple of hours.

Our contribution consists of
\begin{inparaenum}[(a)]
\item the idea of using a UML state machine model to support
  specification of interaction scenarios,
\item the development of a tool (\migtool, Measuring Interaction
  Graphs Tool) that, based on models, transforms scenario
  specifications into interaction paths, and
\item the definition of metrics that provide concise, precise and
  objective measures of the interaction structure entailed by a user
  interface. 
\end{inparaenum}
The examples presented below show that among four web mail
applications, and with respect to a typical usage scenario, Gmail is
the most efficient and flexible UI, with the best ability to recover
from user errors, but only for users that possess a certain level of
proficiency. In fact, Gmail has the largest number of shortest paths
(when users are supposed to make 0 or 1 error); when users make 2 or
more errors the number of paths drops significantly, which indicates
that the error-proneness of the UI has reduced; in the best case,
Gmail features also the shortest paths, requiring 13\% fewer steps;
however, the probability that a user hits an optimal path with Gmail
is 10 times smaller than the best of the other applications, and the
probability that a random walker hits a state that is not due to an
error is 10\% smaller than the best of the other applications.
%
These values suggest that Gmail offers more ways to accomplish tasks
included in the scenario, that comparably more of these ways do not
involve extra steps, that they require fewer steps, and that it might
be more difficult for novice users to exploit the most efficient
ways. Further inspections show that some differences are due to the
slightly different interaction structures adopted for uploading
attachments. Thus, redesigning the methods with which competing
applications deal with attachments could improve their usability. If
Gmail didn't exist yet, by using \migtool\ its designers could obtain
these answers well before developing prototypes and performing
usability studies.

@@

Lostness~\cite{smith1996lostness,otter2000lostness} is a
metric for measuring the degree to which users become lost in the
information space, and therefore considers the notion of
\emph{deviation}. Defined specifically for hypertext systems, lostness
is a user performance measure  which is a function of the number of
visited nodes, the number of different nodes that were visited nodes,
and the number of required nodes. This measure of efficiency is
usually applied to traces of actual users, and is argued to be
suitable for hypertext systems because the predominant task is
browsing information, rather than trying to achieve specific goals. It
is suitable therefore when some of the following three assumptions can
be relaxed: that there is a task to complete, that there is a correct
way to carry it out, that the purpose of the system is to support
users to carry out their tasks.

@@

In \cite{fayollas2014}, hierarchical task models (specified with
HAMSTERS) are used to specify how pilots are expected to interact with
airplane cockpit interfaces in tasks such as changing value to some
flight parameter. The paper in particular discusses how these task
models can be used to specify alternative ways to achieve the same
goal, and analyses the consequences on safety and reliability of the
resulting system. Compared to MIGTool, this approach cannot take
advantage of a model of the system and instead relies only on models
of the tasks. The conseguence is that analysing alternative ways to
achieve the same goal requires developing different models; in
addition, exploring suitability of a given system with respect to
certain tasks is complex because no model of the system is
available. Finally, using \emph{only} hierarchical tasks as a modeling
language limits the richness of the scenarios being analysed, because
it is difficult to take advantage of contexts and rich control flows
that are available with state-transition representations, especially
with UML state diagrams.

A review of usability measuring practices~\cite{hornbaek2006current}
lists, under the headings \emph{measures of efficiency/usage
  patterns}, number of keystrokes, mouse clicks, and visited objects
as possible metrics. Notice that in general the time needed to carry
out individual interaction steps depends on many aspects, such as the
time to decide what goal to achieve, to understand how to carry it
out, to locate the actions to do it, to perform the action, to attend
to the feedback. Layout of the user interface, Fitt's law
considerations, cognitive state of the user, physical and
environmental aspects affect such times. For these reasons we opted,
in \migtool, to only focus on required steps to complete a task, not
on completion time.

A usability analysis method capable to analytically predict task
completion times from a storyboard of the UI is based on using
CogTool~\cite{bellamy2011cogtool}. CogTool relies on the ACT-R
cognitive modeling engine, and allows a designer to setup a
low-fidelity prototype of the UI. After deciding which interaction
modality and which widgets are used to implement actions, the designer
gets an estimation of how long a skilled user would spend on each
step. CogTool takes care of adding extra ``mental'' steps, such as
``think'' steps, before certain patterns of provided steps, according
to the cognitive theory underlying ACT-R. As a result, users of
CogTool obtain the breakdown of the times required by each of the
steps. Our method is less precise: it does not provide expected
completion times. However, with our method, a designer can analyze a
large part of the UI, get information about possible problems in some
areas, and only then devolve more resources in building storyboards
and in making assumptions regarding widgets so that specific execution
paths previously identified can be analyzed with CogTool. In a sense,
the output obtained with our method could be used to make informed
decisions as to what to analyze next with CogTool. Approaches like
\cite{swearngin13} start from the assumption that the user interface
exists already, and that it can \emph{ripped} to reverse engineer a
behavior model. These assumptions are not needed to use \migtool.

Another approach for analysing user actions that has a strong
cognitive background is SNIF-ACT \cite{pirolli2007}. However it has
been conceived and applied to information websites, where the notion
of \emph{information scent} can bear upon interaction. We opted for
not applying such concepts to the interaction with a web application,
where most of the actions are not aimed at information finding.

Markov models, \ie\ directed graphs where edges leaving a vertex are
associated to a probability distribution, were used
in~\cite{thimbleby2001markov}, as a means to perform usability
analysis as early as possible, even before building prototypes of the
UI. Vertices represent states of the UI and edges correspond to user
actions (such as pressing a push-down button).  Probabilities can be
used as a model of user knowledge: equiprobable actions correspond to
a knowledge-free user, whereas when some actions have a very low
probability it means that for that user the action is unlikely to be
executed. Simple mathematical operations on the transition matrix of
the model give the probability that after $n$ steps from a given
initial state the UI is in a given state. With Markov models, by
manipulating probabilities, the designer can plot the number of
required steps as a function of how close the probabilities are to the
designer's ``perfect'' knowledge.  Examples discussed in the paper
cover several devices, ranging from a simple torch (with 4 states), a
microwave cooker (6 states), a mobile phone (152 states). Notice that those are
all push-down devices with a fixed set of buttons. This is obviously
not the case for UIs of information systems, where buttons may change
screen by screen. This makes it more difficult to specify the
transition matrix. 
%
Our approach is based on statecharts, a language that in practice is
more powerful than Markov models, making it easier to specify the UI
behavior, especially in cases where the set of buttons change over
time. While our examples do not make use of probabilities, this is
very easy to cope with (see the Discussion at the end of the paper for
some of the benefits that doing this could bring). Similarly
to~\cite{thimbleby2001markov}, our approach could be used when
conservative results that do not rely on psychological assumptions are
sought.

A discussion of social network analysis metrics applied to interaction
design is provided in~\cite{thimbleby09:sna:eics09}. Once more, a UI
is modeled in terms of directed graphs (vertices are states and edges
are actions), and various centrality metrics are used to draw
conclusions that bear upon usability. For example, centrality measures
(such as Sabidussi, eccentricity, betweenness) can be used to identify
states that are good places to start from to get to other
states. Other metrics, such as edge betweenness, can be used to
identify actions that are important because most of the shortest paths
go through these actions. The paper presents compelling examples of
using this technique to identify shortcomings in the design of
infusion pumps.
%
In our work we automatically generate graphs from statechart models,
and on some of them we computed these metrics. We were not able to draw
sensible conclusions from the values we obtained (for example from the
models presented below). One possible explanation rests on the
different types of models: in our case they reflect the variety and
flexibility with which ``buttons'' can be used in modern web
applications. For infusion pumps the UI is more constrained in how a
task can be carried on, and this difference might reflect on the
usefulness of those metrics.

@@

\section{Generation of interaction traces}
\label{sec:traces}

@@

Because statechart models take advantage of abstraction features and a
rich set of connecting pseudostates, they are not suitable to be
directly processed to compute metrics. For this reason, \migtool\
first \emph{flattens} the model. Flattening is a process often used
when statecharts have to be automatically
processed~\cite{briand2005:test:uml}, and it means to produce a FSA
that is behaviorally equivalent to the original statechart, with no
hierarchy between states and no concurrent regions. In most cases this
leads to an exponential number of states and transitions in the FSA,
but because the process is completely automatic and there is no need
to manually inspect the resulting FSA, this aspect is in many cases
not relevant. Because \migtool\ does not use executable UML,
transition guards are not given semantics and are simply treated as
part of the event of a transition. \migtool\ produces an XML
representation (graphML) of the resulting FSA, the \emph{interaction
  graph}. It is a directed multigraph\footnote{A directed multigraph
  is a directed graph such that there are 2 or more edges that have
  the same end points. Cycles are paths that include 2 or more
  occurrences of the same vertex. Loops are edges that start and end
  on the same vertex. ``Geodesic path'' is a synonymous term with
  ``shortest path''.}, potentially with cycles and loops, with edges
that are labeled with the name of the corresponding action. In the
following we will be using as synonymous the terms state/vertex, and
action/edge.

@@

Because in all but the most trivial interaction graphs there are loops
or cycles, the set of possible interaction traces is infinite. For
this reason, scenarios need to be defined and used as constraints on
the possible execution traces that can be generated.  Users of
\migtool\ define interaction scenarios by specifying key steps (called
bridge sets) that users are expected to go through; we call
this process \emph{grounding usage scenarios on models}. A \emph{bridge set}
is a subset of the edges of the interaction graph; a well formed
bridge set is a non-empty set (an empty bridge set would make the
scenario unviable). In general a specification of a scenario includes
an initial state and a non-empty sequence bridge sets. For example, to
specify a scenario for replying to an email message, assuming an
initial state that corresponds to a user interface that displays all
messaes in the inbox, one could select all the edges associated to the
action \texttt{reply} (bridge set 1), followed by edges labeled with
\texttt{typeBody} (bridge set 2), followed by edges labeled with
\texttt{send} (bridge set 3).  In this way scenarios with cycles can
be formulated (\eg, reply twice to two messages). A \emph{stage} of a
scenario comprises two consecutive bridge sets.
%
To cope with multi edges, the interaction graph is \emph{simplified}:
all edges between a   pair of vertices are merged into a single one,
whose label includes the original ones.

@@

Quite expressive languages can be conceived for grounding scenarios
(\eg, regular expressions on sequences of action names). But such
expressivity bonus needs to be balanced with computational
tractability: even models with a dozen states might correspond to interaction
graphs with several hundred states and several thousand edges, leading
to an enormous number of possible traces to filter even for scenarios
with just a few stages.

To cope with this we implemented a trace searching algorithm that
processes each of the stages sequentially, starting from the
initial state. Given a bridge set $B_i$, the algorithm does a
breadth-first search of all the geodesic paths that connect
\emph{each} of the ending vertices of edges in $B_i$ with \emph{some}
of the starting vertices of edges in $B_{i+1}$. If some bridge in
$B_{i+1}$ cannot be reached, then it is dropped from further
searches. \migtool\ creates a new graph from the geodesic paths found
for each stage, and then joins these graphs so that geodesic paths
found for stage $i$ are joined with those of ${i+1}$. These global
paths, connecting the initial state to reachable bridges of the
last bridge set of the scenario, are called \emph{execution traces}.

Notice that a scenario specifies only the desired occurrences of
actions, not all the necessary ones. For example, if the model
prescribes that in order to \texttt{view(message)} while reading
another one, one has to \texttt{goBack} to the inbox first, a scenario
specifying two consecutive \texttt{view(message)} would lead to traces
that include also the \texttt{goBack} action, even if that step is not
explicitly specified in the scenario. It is the task of \migtool\ to
unfold paths in the graph and search all the geodesic paths that
connect the desired user actions.

@@

\migtool\ is used to process a model and a scenario
and to generate execution traces of detour order = 0,
..., H. 
For each stage, up to a maximum detour order $H$, the search algorithm
creates traces with detour order $k+1$ by collecting the set $D_k$ of states
with order $0$ up to $k$, and by identifying the neighbors $N_k$
of $D_k$ ($N_k$ is the set of states  not included in $D_k$ that are
connected through an edge to some state in $D_k$). Edges connecting
$D_k$ with $N_k$ represent deviations that users might follow, and
geodesic paths from $N_k$ to $D_k$ constitute recovery paths that
users might follow to complete the scenario from states in
$N_k$. These deviations and recovery paths are joined and constitute
the traces of order $k+1$.  It could happen that
for some state in $N_k$ there is no path leading to any vertex in
$D_k$: in such a case the deviation is a dead end that prevents
the user to complete the scenario.

@@

Figure~\ref{fig:model} shows part of the model of Gmail. At some
point, a user is viewing all the conversations of the inbox (state
\texttt{viewingConversations}); available actions include moving to
the next or previous block of conversations, refreshing the list, or
opening a specific conversation (transition
\texttt{open(conversation)}). This transition is assumed to occur when
the user clicks on any one of the visible conversations - it is if
there is a collection of transitions, each one for a different
conversation that can be opened; the parameter \texttt{conversation}
is bound to any occurrence of a conversation. Such transitions lead to the state
\texttt{viewingAConversation}, where the behavior of the system is
defined by a more detailed state machine. By default the user is
viewing a conversation, but by performing the \texttt{reply} action
the UI moves to a state called \texttt{replying}, where the body of
the reply can be typed, the subject can be changed, or another
addressee can be added.

@@

Manual inspection of the shortest paths indicates that one reason of
the greated potential efficiency offered by Gmail is due to the
fact that users can start composing a new message while reading a
conversation, whereas in other applications an explicit ``close action'' of
the reading activity has to be performed. Another reason relies on the
more streamlined process to attach a file: in Gmail one needs to
select the file(s) and they are automatically uploaded, whereas in
other applications one has to explicitly perform the uploading step
after selecting them.

>>>>

\documentclass{sig-alternate-05-2015}

@@

% Load basic packages
\usepackage{balance}  % to better equalize the last page
\usepackage{graphics} % for EPS, load graphicx instead 
%\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage{times}    % comment if you want LaTeX's default font
\usepackage[pdftex]{hyperref}
% \usepackage{url}      % llt: nicely formatted URLs
\usepackage{color}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{ccicons}
\usepackage{todonotes}
\newcommand{\mytodo}[1]{\todo[inline]{#1}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\migtool}{MIG\-tool}
\usepackage{soul}
\usepackage{paralist}
\usepackage{listings}

@@

% \author{Giorgio Brajnik\inst{1,2} \and Simon Harper\inst{1}}
% \institute{\textsuperscript{1}Computer Science School\\
% University of Manchester, UK\\
% \email{simon.harper@manchester.ac.uk}\\
% and\\
% \inst{2}Dipartimento di Matematica e Informatica\\
% Universit\`a di Udine, Italy\\
% \email{brajnik@uniud.it}
% }
\CopyrightYear{2016}
\setcopyright{acmlicensed}
\conferenceinfo{EICS'16,}{June 21 - 24, 2016, Brussels, Belgium}
\isbn{978-1-4503-4322-0/16/06}\acmPrice{\$15.00}
\doi{http://dx.doi.org/10.1145/2933242.2933246}

@@

\keywords{User Interfaces; Evaluation; Statecharts: UML; Metrics; Usability.}

@@

\newpage
\section{Introduction}

We present an approach that allows a designer to assess interaction
design qualities such as efficiency, error-proneness and
recovery from errors. Key importance is given to the ability to
\begin{inparaenum}[(a)]
\item understand how supportive a user interface (UI) is with respect
  to user efficiency;
\item understand how prone the UI is to user navigation
  errors;
\item understand how recoverable the UI is from those errors; and,
\item perform objective quantitative comparisons of different designs
  to support construction and engineering of interactive systems.
\end{inparaenum}
All this can be done before developing prototypes of the UI, similarly
to what is claimed and done in~\cite{silva14}: ``... the analysis of solutions in
the early phases of development is crucial as their results can then
be iteratively injected in the development cycle''.

@@

Developing good UIs for web or mobile applications is a
complex and expensive endeavor. One reason is the combination of
devices, interaction modalities and workflows that need to be
supported.
%
Adoption of usage-centered development practices and following
established design principles~\cite{constantine99} is a sound way to
tackle the problem.  In particular very effective techniques are
prototyping, to explore part of the five-dimensional fidelity
space~\cite{mccurdy06,buxton07}, paired with usability investigations,
based on user testing or heuristic
evaluations~\cite{preece02,rubin08}.

@@

Even worse, several questions cannot be easily
answered. Given one or more potential designs and some usage
scenarios, interesting questions include: ``How many different
routes can be followed by the user to carry out the scenario?'',
``Which are the shortest ones?'', ``If a user makes a mistake, would
he or she be able to recover?'', ``How many steps would the recovery
require?''. When designing and evaluating
embedded UIs (such as when dealing with plane's
cockpits~\cite{conversy07}), other relevant questions include
``How would the above properties change if we add a certain a
widget?'', or ``... if we replace a widget with another?''.  At the
moment, these straightforward questions are quite complicated to
answer. In fact, they require developing  prototypes, inspecting
them, manually tracking which screens and widgets are used at which
stage, followed by a systematic manual analysis.

This should not be the case, however, because answers to these
questions could be automatically computed. Such a system could
provide important insights to a designer, and support assessments of
potential user flexibility, user efficiency, error proneness, ability
to recover, compactness and consistency of a design.

Our approach is based on UML state machine models of UIs which are
automatically processed to produce \emph{interaction graphs}. These
graphs are then used to specify interaction scenarios and to unfold
the possible interaction paths (called \emph{execution traces}) that
are compatible with the specific scenarios being considered. Traces
are processed to produce a dashboard with different results. Except
for development of models and specification of the desired scenarios,
which have to be done manually, the other steps are totally automatic;
models of a design (such as the ones shown below) can be developed in
a matter of a couple of hours.

Our contribution consists of
\begin{inparaenum}[(a)]
\item the idea of using a UML state machine model to support
  specification of interaction scenarios,
\item the development of a tool (\migtool, Measuring Interaction
  Graphs Tool) that, based on models, transforms scenario
  specifications into interaction paths, and
\item the definition of metrics that provide concise, precise and
  objective measures of the interaction structure entailed by a user
  interface. 
\end{inparaenum}
The examples presented below show that among four web mail
applications, and with respect to a typical usage scenario, Gmail is
the most efficient and flexible UI, with the best ability to recover
from user errors, but only for users that possess a certain level of
proficiency. In fact, Gmail has the largest number of shortest paths
(when users are supposed to make 0 or 1 error); when users make 2 or
more errors the number of paths drops significantly, which indicates
that the error-proneness of the UI has reduced; in the best case,
Gmail features also the shortest paths, requiring 13\% fewer steps
than other applications; however, the probability that a user hits an
optimal path with Gmail is 10 times smaller than the best of the other
applications, and the probability that a random walker hits a state
that is not due to an error is 10\% smaller than the best of the other
applications.
%
These values suggest that Gmail offers more ways to accomplish tasks
included in the scenario, that comparably more of these ways do not
involve extra steps, that they require fewer steps, and that it might
be more difficult for novice users to exploit the most efficient
ways. Further inspections show that some differences are due to the
slightly different interaction structures adopted for uploading
attachments. Thus, redesigning the methods with which competing
applications deal with attachments could improve their usability. If
Gmail didn't exist yet, by using \migtool\ its designers could obtain
these answers well before developing prototypes and performing
usability studies.

@@

On the other hand the literature on metrics is also large.
Lostness~\cite{smith1996lostness,otter2000lostness} is a metric for
measuring the degree to which users become lost in the information
space, and it considers the notion of \emph{deviation}. Defined
specifically for hypertext systems, lostness is a user performance
measure which is a function of the number of visited nodes, the number
of different nodes that were visited, and the number of required
nodes. This measure of efficiency is usually applied to traces of
actual users, and is argued to be suitable for hypertext systems
because the predominant task is browsing information, rather than
trying to achieve specific goals. It is suitable therefore when some
of the following three assumptions can be relaxed: that there is a
task to complete, that there is a correct way to carry it out, that
the purpose of the system is to support users in carrying out their
tasks.

@@

In \cite{fayollas2014}, hierarchical task models (specified with
HAMSTERS) are used to specify how pilots are expected to interact with
airplane cockpit interfaces in tasks such as changing value to some
flight parameter. The paper in particular discusses how these task
models can be used to specify alternative ways to achieve the same
goal, and analyses the consequences on safety and reliability of the
resulting system. Compared to MIGTool, this approach cannot take
advantage of a model of the system and instead relies only on models
of the tasks. The consequence is that analyzing alternative ways to
achieve the same goal requires developing different models; in
addition, exploring suitability of a given system with respect to
certain tasks is complex because no model of the system is
available. Finally, using \emph{only} hierarchical tasks as a modeling
language limits the richness of the scenarios being analyzed, because
it is difficult to take advantage of contexts and rich control flows
that are available with state-transition representations, especially
with UML state diagrams.

A review of usability measuring practices~\cite{hornbaek2006current}
lists, under the headings \emph{measures of efficiency/usage
  patterns}, number of keystrokes, mouse clicks, and visited objects
as possible metrics. Of course also the time needed to carry
out individual interaction steps counts and it depends on many aspects, such as the
time to decide what goal to achieve, to understand how to carry it
out, to locate the actions to do it, to perform the action, to attend
to the feedback. Layout of the user interface, Fitt's law
considerations, cognitive state of the user, physical and
environmental aspects affect such times. For these reasons we opted,
in \migtool, to only focus on required steps to complete a task, not
on completion time.

A usability analysis method capable to analytically predict task
completion times from a storyboard of the UI is based on using
CogTool~\cite{bellamy2011cogtool}. CogTool relies on the ACT-R
cognitive modeling engine, and allows a designer to setup a
low-fidelity prototype of the UI. After deciding which interaction
modality and which widgets are used to implement actions, the designer
gets an estimation of how long a skilled user would spend on each
step. CogTool takes care of adding extra ``mental'' steps before
certain patterns of provided steps, according to the cognitive theory
underlying ACT-R. As a result, users of CogTool obtain the breakdown
of the times required by the task. Our method is less precise: it does
not provide expected completion times. However, with our method, a
designer can analyze a large part of the UI, get information about
possible problems in some areas, and only then devolve more resources
in building storyboards and in making assumptions regarding widgets so
that specific execution paths previously identified can be analyzed
with CogTool. In a sense, the output obtained with our method could be
used to make informed decisions as to what to analyze next with
CogTool. Other approaches, like \cite{swearngin13}, start from the assumption
that the user interface exists already, and that it can \emph{ripped}
to reverse engineer a behavior model. These assumptions are not needed
when using \migtool.

Another approach for analyzing user actions that has a strong
cognitive background is SNIF-ACT \cite{pirolli2007}. However it has
been conceived and applied to information websites, where the notion
of \emph{information scent}  bears upon interaction. We opted for
not applying such concepts to the interaction with a web application,
where most of the actions are not aimed at information finding.

Markov models, \ie\ directed graphs where edges leaving a vertex are
associated to a probability distribution, were used
in~\cite{thimbleby2001markov} as a means to perform usability analysis
as early as possible, even before building prototypes of the
UI. Vertices represent states of the UI and edges correspond to user
actions (such as pressing a button).  Probabilities can be used as a
model of user knowledge: equiprobable actions correspond to a
knowledge-free user, whereas when some actions have a very low
probability it means that for that user the action is unlikely to be
executed. Simple mathematical operations on the transition matrix of
the model give the probability that after $n$ steps from a given
initial state the UI is in a given state. With Markov models, by
manipulating probabilities, the designer can plot the number of
required steps as a function of how close the probabilities are to the
designer's ``perfect'' knowledge.  Examples discussed in the paper
cover several devices, ranging from a simple torch (with 4 states), a
microwave cooker (6 states), a mobile phone (152 states). Notice that
those are all push-down devices with a fixed set of buttons. This is
obviously not the case for UIs of information systems, where buttons
may change screen by screen and there could also be an arbitrary
number of links. This makes it more difficult to specify the
transition matrix.
%
Our approach is based on statecharts, a language that in practice is
more powerful than Markov models, making it easier to specify the UI
behavior, especially in cases where the set of buttons change over
time. While our examples do not make use of probabilities, this is
very easy to cope with (see the Discussion at the end of the paper for
some of the benefits that doing this could bring). Similarly
to~\cite{thimbleby2001markov}, our approach could be used when
conservative results that do not rely on psychological assumptions are
sought.

A discussion of social network analysis metrics applied to interaction
design is provided in~\cite{thimbleby09:sna:eics09}. Once more, a UI
is modeled in terms of directed graphs (vertices are states and edges
are actions), and various centrality metrics are used to draw
conclusions that bear upon usability. For example, centrality measures
(such as Sabidussi, eccentricity, betweenness) can be used to identify
states that are good places to start from to get to other
states. Other metrics, such as edge betweenness, can be used to
identify actions that are important because most of the shortest paths
go through these actions. The paper presents compelling examples of
using this technique to identify shortcomings in the design of
infusion pumps.
%
In our work we automatically generate graphs from statechart models,
and on some of them we computed these metrics. In case studies
presented below we were not able to draw sensible conclusions from the
values we obtained (for example from the models presented below). One
possible explanation rests on the different types of models: in our
case they reflect the variety and flexibility with which ``buttons''
can be used in modern web applications. For infusion pumps the UI is
more constrained in how a task can be carried on, and this difference
might reflect on the usefulness of those metrics.

@@

\section{Generation of traces}
\label{sec:traces}

@@

Because statechart models take advantage of abstraction features and a
rich set of connecting pseudostates, they are not suitable to be
directly processed to compute metrics. For this reason, \migtool\
first \emph{flattens} the model. Flattening is a process often used
when statecharts have to be automatically
processed~\cite{briand2005:test:uml}, and it means to produce a FSA
that is behaviorally equivalent to the original statechart, with no
hierarchy between states and no concurrent regions. In most cases this
leads to an exponential number of states and transitions in the FSA,
but because the process is completely automatic and there is no need
to manually inspect the resulting FSA, this aspect is in many cases
not relevant. Because \migtool\ does not use executable UML,
transition guards are not given semantics and are simply treated as
part of the event label of a transition. \migtool\ produces an XML
representation (graphML) of the resulting FSA, the \emph{interaction
  graph}. It is a directed multigraph\footnote{A directed multigraph
  is a directed graph such that there are 2 or more edges that have
  the same end points. Cycles are paths that include 2 or more
  occurrences of the same vertex. Loops are edges that start and end
  on the same vertex. ``Geodesic path'' is a synonymous term with
  ``shortest path''.}, potentially with cycles and loops, with edges
labeled with the name of the corresponding action. In the following we
will be using as synonymous the terms state/vertex, action/edge and
trace/path.

@@

Because in all but the most trivial interaction graphs there are loops
or cycles, the set of possible interaction traces is infinite. For
this reason, scenarios need to be defined and used as constraints on
the possible execution traces that can be generated.  Users of
\migtool\ define interaction scenarios by specifying key steps (called
bridge sets) that users are expected to go through; we call
this process \emph{grounding usage scenarios on models}. A \emph{bridge set}
is a subset of the edges of the interaction graph; a well formed
bridge set is a non-empty set (an empty bridge set would make the
scenario unviable). In general a specification of a scenario includes
an initial state and a non-empty sequence of bridge sets. For example, to
specify a scenario for replying to an email message, assuming an
initial state that corresponds to a user interface that displays all
messages in the inbox, one could select all the edges associated to the
action \texttt{reply} (bridge set 1), followed by edges labeled with
\texttt{typeBody} (bridge set 2), followed by edges labeled with
\texttt{send} (bridge set 3).  In this way scenarios with cycles can
be formulated (\eg, reply twice to two messages). A \emph{stage} of a
scenario comprises two consecutive bridge sets.
%
To cope with multi edges, the interaction graph is \emph{simplified}:
all edges between a   pair of vertices are merged into a single one,
whose label includes the original ones.

@@

Quite expressive languages can be conceived for grounding scenarios
(\eg, regular expressions on sequences of action labels). But such
expressivity bonus needs to be balanced with computational
tractability: even models with a dozen states might correspond to interaction
graphs with several hundred states and several thousand edges, leading
to an enormous number of possible traces to filter even for scenarios
with just a few stages.

To cope with this we implemented a trace searching algorithm that
processes each of the stages sequentially, starting from the initial
state. Given a stage $i$ and a bridge set $B_i$, the algorithm does a
breadth-first search of all the geodesic paths that connect
\emph{each} of the ending vertices of edges in $B_i$ with \emph{some}
of the starting vertices of edges in $B_{i+1}$. If some bridge in
$B_{i+1}$ cannot be reached, then it is dropped from further
searches. \migtool\ creates a new graph from the geodesic paths found
for each stage, and then joins these graphs so that geodesic paths
found for stage $i$ are joined with those of ${i+1}$. These global
paths, connecting the initial state to reachable bridges of the last
bridge set of the scenario, are called \emph{execution traces}.

Notice that a scenario specifies only the desired occurrences of
actions, not all the necessary ones. For example, if the model
prescribes that in order to perform action \texttt{view(message)}
while reading another message one has to \texttt{goBack} to the inbox
first, a scenario specifying two consecutive \texttt{view(message)}
would lead to traces that include also the \texttt{goBack} action,
even if that step is not explicitly specified in the scenario. It is
the task of \migtool\ to unfold paths in the graph and search all the
geodesic paths that connect the desired user actions.

@@

\migtool\ is used to process a model and a scenario and to generate
execution traces of detour order = 0, ..., H.  For each stage, up to a
maximum detour order $H$, the search algorithm creates traces with
detour order $k+1$ by collecting the set $D_k$ of states with order
$0$ up to $k$, and by identifying the neighbors $N_k$ of $D_k$ ($N_k$
is the set of states not included in $D_k$ that can be reached through
an edge from some state in $D_k$). Edges connecting $D_k$ with $N_k$
represent deviations that users might follow, and geodesic paths from
$N_k$ to $D_k$ constitute recovery paths that users might follow to
complete the scenario from states in $N_k$. These deviations and
recovery paths are joined and constitute the traces of order $k+1$.
It could happen that for some state in $N_k$ there is no path leading
to any vertex in $D_k$: in such a case the deviation is a dead end
that prevents the user to complete the scenario.

@@

Figure~\ref{fig:model} shows part of the model of Gmail. At some
point, a user is viewing all the conversations of the inbox (state
\texttt{viewing\-Conversations}); available actions include moving to
the next or previous block of conversations, refreshing the list, or
opening a specific conversation (transition
\texttt{open(conversation)}). This transition is assumed to occur when
the user clicks on any one of the visible conversations - it is as if
there is a collection of transitions, each one for a different
conversation that can be opened; the parameter \texttt{conversation}
is conceptually bound to any occurrence of a conversation. Such
transitions lead to the state \texttt{viewingAConversation}, where the
behavior of the system is defined by a more detailed state machine. By
default the user is viewing a conversation, but by performing the
\texttt{reply} action the UI moves to a state called
\texttt{replying}, where the body of the reply can be typed, the
subject can be changed, or another addressee can be added.

@@

Manual inspection of the shortest paths indicates that one reason for
the greater potential efficiency offered by Gmail is due to the
fact that users can start composing a new message while reading a
conversation, whereas in other applications an explicit ``close action'' of
the reading activity has to be performed. Another reason relies on the
more streamlined process to attach a file: in Gmail one needs to
select the file(s) and they are automatically uploaded, whereas in
other applications one has to explicitly perform the uploading step
after selecting them.
